---
title: "RNAseq Differential Expression"
author: "Annat Haber"
date: "5/1/2021"
output:
  html_document:
    toc: true
    code_folding: hide
    df_print: "kable"
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message=FALSE, warning=FALSE, cache = TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(synapser)
  library(DESeq2)
  library(tidyverse)
 })
```

## Input
1. Read sample metadata and raw counts.  
2. Switch genotypes for two samples based on [the QC](RNAseq_1_Processing_QC):  
    34158 from TT to CT  
    34163 from CT to TT  
3. Sort everything by sample ID.  
```{r input}
# read sample info
meta <- readRDS("output/RNAseq/1_Processing_QC/samplesheet_meta.RDS")
meta$genotype[meta$sample_id=="34158"] <- "CT"
meta$genotype[meta$sample_id=="34163"] <- "TT"
meta <- meta %>%
  arrange(genotype, sex, age)

# read raw counts
ctsfile <- "output/RNAseq/1_Processing_QC/nf_rnaseq_output/star_rsem/rsem.merged.gene_counts.tsv"
counts <- read_tsv(ctsfile) %>%
  dplyr::select(-"transcript_id(s)") %>%
  column_to_rownames("gene_id") %>%
  round() %>%
  as.matrix()

# replace pipeline sample id with original sample id
counts <- counts[,meta$samplesheet]
colnames(counts) <- meta$sample_id
```

## Sample exploration
Using variance stabilizing transformation for visualization.
There is great separation between male and female. In females there is separation of age, but not in males. There is no separation at all between genotypes.
```{r plotPCA}
# Modifying plotPCA function to include PC3
plotPCA = function(object, ntop=500, intgroup)
{
# calculate the variance for each gene
  rv <- rowVars(assay(object))
  # select the ntop genes by variance
  select <- order(rv, decreasing=TRUE)[seq_len(min(ntop, length(rv)))]
  # perform a PCA on the data in assay(x) for the selected genes
  pca <- prcomp(t(assay(object)[select,]))
  # the contribution to the total variance for each component
  percentVar <- pca$sdev^2 / sum( pca$sdev^2 )
  
  intgroup.df <- as.data.frame(colData(object)[, intgroup, drop=FALSE])
  # add the intgroup factors together to create a new grouping factor
  group <- if (length(intgroup) > 1) {
    factor(apply( intgroup.df, 1, paste, collapse=":"))
  } else {
    colData(object)[[intgroup]]
  }
  # assemble the data for the plot
  d <- data.frame(PC1=pca$x[,1], PC2=pca$x[,2], PC3=pca$x[,3], group=group, intgroup.df, name=colnames(object))
  attr(d, "percentVar") <- percentVar[1:3]
  return(d)
}
```

```{r sample_exp, out.width=c('50%', '50%'), fig.show='hold', out.extra='style="background-color: #9ecff7; padding:1px; display: inline-block;"'}
coldata <- meta %>%
  dplyr::select(sample_id, genotype, sex, age) %>%
  mutate(age=as.numeric(age)) %>%
  column_to_rownames("sample_id")

dds <- DESeqDataSetFromMatrix(counts, coldata, design = ~ genotype)
dds <- dds[rowSums(counts(dds)) > 10,]
dds$genotype <- relevel(dds$genotype, ref = "CC")
vsd <- vst(dds, blind=FALSE)

pcaData <- plotPCA(vsd, intgroup=c("genotype", "sex", "age"))
pcaData$age <- factor(pcaData$age)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x=PC1,y=PC2)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(1.5,3)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
ggplot(pcaData, aes(x=PC1,y=PC3)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(1.5,3)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC3: ",percentVar[3],"% variance")) + 
  coord_fixed()

rm(dds)
```

## Differential expression {.tabset .tabset-fade .tabset-pills}
### Variant vs WT {.tabset .tabset-fade .tabset-pills}
For each combination of sex (M,F) and age (6,18) we compare CT to CC and TT to CC.  
Genes that have less than 10x reads total are filtered out within each comparison.  
Genes with p-adj < 0.1 are in blue.
```{r deseq_obj_gt}
grp <- meta %>%
  filter(genotype!="CC") %>%
  distinct(genotype, sex, age)

de_resL <- list()
sig.genes_gt <- c()

for (i in 1:nrow(grp)) {
  #print(i)
  # get genotype, sex, and age for this comparison
  gt <- grp$genotype[i]
  sx <- grp$sex[i]
  ag <- grp$age[i]
  gr <- paste(gt, sx, ag, sep="_")
  
  # get sample_id and grouping factor (genotype) for this comparison
  coldata <- meta %>%
    filter(genotype %in% c("CC",gt) & sex == sx & age == ag) %>%
    dplyr::select(genotype, sample_id) %>%
    column_to_rownames("sample_id")

  cts <- counts[, rownames(coldata)]

  # DESeq object and analysis
  dds <- DESeqDataSetFromMatrix(cts, coldata, design = ~ genotype)
  dds <- dds[rowSums(counts(dds)) > 10,]
  dds$genotype <- relevel(dds$genotype, ref = "CC")
  dds <- DESeq(dds) 
 
  # extract DEseq results
  res <- results(dds)
  res <- res[order(res),]
  de_resL[[gr]] <- res
  
  # get genes significant at padj < 0.1
  sig.g <- as.data.frame(res[which(res$padj<0.1),])
  if (nrow(sig.g) > 0) {
    sig.g <- cbind(comparison=gr, sig.g) %>%
      arrange(desc(padj)) %>%
      rownames_to_column(var='ensembl_gene_id')
    # add gene symbol
    ensembl <- biomaRt::useMart("ensembl", dataset="mmusculus_gene_ensembl")
    mgi_symbol <- biomaRt::getBM(attributes=c('ensembl_gene_id','external_gene_name'),
             filters = 'ensembl_gene_id',
             values = sig.g$ensembl_gene_id,
             mart = ensembl) %>%
      dplyr::rename("mgi_symbol"="external_gene_name")
    sig.g <- left_join(sig.g, mgi_symbol, by='ensembl_gene_id')
    sig.genes_gt <- rbind(sig.genes_gt, sig.g)
  }
  
  rm(gt, sx, ag, coldata, cts, dds, res, sig.g)
}

```

```{r deseq_res_gt, results='asis'}
for(gr in names(de_resL)){
  cat("#### ", gr ,"\n")
  res <- de_resL[[gr]]
  cat(paste0("Comparing ", gr, " to ", str_replace(gr, "CT|TT", "CC")), "\n")
  cat(plotMA(res), "\n")
  #cat("\n", "significant genes:\n", rownames(res)[which(res$padj<0.1)])
  cat('\n\n')
}

de_resL_gt <- de_resL
rm(de_resL)
```

### Old vs Young {.tabset .tabset-fade .tabset-pills}

For each combination of sex (M,F) and genotype (CC, CT, TT) we compare old to young samples (18 vs 6 months).  
Genes that have less than 10x reads total are filtered out within each comparison.  
Genes with p-adj < 0.1 are in blue.
```{r deseq_obj_age}
grp <- meta %>%
  filter(age!="6") %>%
  distinct(genotype, sex, age)

de_resL <- list()
sig.genes_age <- c()

for (i in 1:nrow(grp)) {
  #print(i)
  # get genotype, sex, and age for this comparison
  gt <- grp$genotype[i]
  sx <- grp$sex[i]
  ag <- grp$age[i]
  gr <- paste(gt, sx, ag, sep="_")
  
  # get sample_id and grouping factor (age) for this comparison
  coldata <- meta %>%
    filter(sex == sx & genotype == gt) %>%
    dplyr::select(age, sample_id) %>%
    column_to_rownames("sample_id")

  cts <- counts[, rownames(coldata)]

  # DESeq object and analysis
  dds <- DESeqDataSetFromMatrix(cts, coldata, design = ~ age)
  dds <- dds[rowSums(counts(dds)) > 10,]
  dds$age <- relevel(dds$age, ref = "6")
  dds <- DESeq(dds) # differential expression analysis
 
  # extract DEseq results
  res <- results(dds)
  res <- res[order(res),]
  de_resL[[gr]] <- res
  
  # get genes significant at padj < 0.1
  sig.g <- as.data.frame(res[which(res$padj<0.1),])
  if (nrow(sig.g) > 0) {
    sig.g <- cbind(comparison=gr, sig.g) %>%
      arrange(desc(padj)) %>%
      rownames_to_column(var='ensembl_gene_id')
    # add gene symbol
    ensembl <- biomaRt::useMart("ensembl", dataset="mmusculus_gene_ensembl")
    mgi_symbol <- biomaRt::getBM(attributes=c('ensembl_gene_id','external_gene_name'),
             filters = 'ensembl_gene_id',
             values = sig.g$ensembl_gene_id,
             mart = ensembl) %>%
      dplyr::rename("mgi_symbol"="external_gene_name")
    sig.g <- left_join(sig.g, mgi_symbol, by='ensembl_gene_id')
    sig.genes_age <- rbind(sig.genes_age, sig.g)
  }

  rm(gt, sx, ag, coldata, cts, dds, res, sig.g)
}
```

```{r deseq_res_age, results='asis'}
for(gr in names(de_resL)){
  cat("#### ", gr ,"\n")
  res <- de_resL[[gr]]
  cat(paste0("Comparing ", gr, " to ", str_replace(gr, "18", "6")), "\n")
  cat(plotMA(res), "\n")
  #cat("\n", "significant genes:\n", rownames(res)[which(res$padj<0.1)])
  cat('\n\n')
}

de_resL_age <- de_resL
rm(de_resL)
```

## Output
```{r}
write_csv(sig.genes_gt,
            file="output/RNAseq/2_Differential_Expression/sig_genes_padj01_variantVSwt.csv")

write_csv(sig.genes_age, 
            file="output/RNAseq/2_Differential_Expression/sig_genes_padj01_18vs6.csv")

saveRDS(de_resL_gt, file="output/RNAseq/2_Differential_Expression/sig.genes_gt.RDS")
saveRDS(de_resL_age, file="output/RNAseq/2_Differential_Expression/sig.genes_age.RDS")
```
