---
title: "MTHFR RRBS Exploratory Analysis"
author: "Annat Haber"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(cache=TRUE, cache.extra = R.version, 
                      #cache.path = "output/RRBS/2_Exploratory_Analysis/",
                      message=FALSE, warning=FALSE)
```

```{r libraries}
rm(list=ls())

suppressPackageStartupMessages({
  library(synapser)
  #library(methylKit)
  library(tidyverse)
  library(betareg)
  library(emmeans)
  library(ComplexHeatmap)
})
```

## Input
Input is percent methylation in each site for each sample (samples x sites matrix), including only sites within CpG context that are covered in all samples with at least 10x coverage. 
```{r input}
site_meth <- readRDS(file = "output/RRBS/1_Processing_QC/persite_meth.RDS")
# meta
meta <- readRDS(file = "output/RRBS/sample_meta.RDS")

hmc <- read_tsv("data/homocysteine.txt", col_types = cols("c","n"))

meta <- left_join(meta, hmc, by="tag_id")
```

## Pattern of Methylation {.tabset .tabset-fade .tabset-pills}

### With Outliers
#### PCA
```{r, pattern_withOutliers}
pc <- prcomp(site_meth, center = TRUE)
plot(pc)

df <- left_join(dplyr::select(meta, age, sex, genotype, sample_id),
                as_tibble(pc$x[,1:2], rownames = "sample_id"),
                by="sample_id")

outliers <- df %>% filter(abs(PC1) > 2000 | abs(PC2) > 2000)
  
ggplot(df, aes(x=PC1,y=PC2)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(3,1.5)) +
  geom_text(data=outliers, aes(label=sample_id), vjust = "inward",  hjust = "inward")

rm(df)
```

```{r, bySite_withOutliers, eval=FALSE, include=FALSE}

cols = list(sex = c("F" = "pink", "M" = "lightblue"),
           age = c("6" = "blue", "18" = "darkblue"),
           genotype = c("CC" = "yellow", "CT" = "orange", "TT" = "red"))
# Create the heatmap annotation
ha <- HeatmapAnnotation(sex = meta$sex, age = meta$age, genotype = meta$genotype, col=cols)

Heatmap(t(site_meth),
        show_row_dend = FALSE,
        show_row_names = FALSE,
        heatmap_legend_param = list(title = "Methlation %"),
        #cluster_rows = FALSE, 
        #clustering_method_columns="ward.D2", 
        #col = hcl.colors(12, "YlOrRd", rev = TRUE),
        top_annotation = ha
        )
```

### Without Outliers
```{r rm_oul}
site_meth.wou <- site_meth[-which(rownames(site_meth) %in% outliers$sample_id),]
meta.wou <- filter(meta, !sample_id %in% outliers$sample_id)
```

#### PCA
```{r, pattern__noOutliers}
pc <- prcomp(site_meth.wou, center = TRUE)
plot(pc)

left_join(select(meta.wou, age, sex, genotype, sample_id),
                as_tibble(pc$x[,1:2], rownames = "sample_id"),
                by="sample_id") %>%
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(3,1.5)) 
#+
#  geom_text(data=outliers, aes(label=sample_id), vjust = -0.5,  hjust = -0.5)
```
  
 
```{r, bySite_noOutliers, include=FALSE, eval=FALSE}
cols = list(sex = c("F" = "pink", "M" = "lightblue"),
           age = c("6" = "blue", "18" = "darkblue"),
           genotype = c("CC" = "yellow", "CT" = "orange", "TT" = "red"))
# Create the heatmap annotation
ha <- HeatmapAnnotation(sex = meta.wou$sex, age = meta.wou$age, genotype = meta.wou$genotype, col=cols)

Heatmap(t(site_meth.wou),
        show_row_dend = FALSE,
        show_row_names = FALSE,
        heatmap_legend_param = list(title = "Methlation %"),
        #cluster_rows = FALSE, 
        #clustering_method_columns="ward.D2", 
        col = hcl.colors(12, "YlOrRd", rev = TRUE),
        top_annotation = ha
        )
```

## Global methylation {.tabset .tabset-fade .tabset-pills}
### Counts
Calculating global methylation as the fraction of sites determined as methylated by bismark for each sample.
```{r global_meth_counts}
global_meth_counts <- read_tsv("output/RRBS/1_Processing_QC/nf_methylseq_output/bismark_summary/bismark_summary_report.txt") %>%
  tibble(.name_repair = "universal") %>%
  mutate(sample_id = substr(File, 1, 5),
         global_meth_counts = Methylated.CpGs / (Methylated.CpGs + Unmethylated.CpGs) ) %>%
  select(global_meth_counts, sample_id)

meta.gm <-  left_join(meta, global_meth_counts, by="sample_id")

```

Using Beta regression to test differences in global methylation as a function of sex, age, and genotype. None of the factors and interactions is significant.
```{r, global_meth_counts_betareg}

br.df <- meta.gm %>%
  mutate(genotype = recode(genotype, "CC"=0, "CT"=1, "TT"=2),
         sex = recode(sex, "F"=0, "M"=1),
         age = recode(age, "6"=0, "18"=1))
   
model = betareg(global_meth_counts ~ genotype + age + sex, data = br.df)
emmeans::joint_tests(model)

ggplot(meta.gm, aes(x=genotype, y=global_meth_counts, color=age)) +
  geom_point() +
  facet_wrap(~ sex)

```

### Mean Percent
Calculating global methylation as the mean percent methylation for each sample, averaged across sites.
```{r global_meth_percent}
global_meth_perc <- data.frame(global_meth_perc=rowMeans(site_meth)/100, sample_id=rownames(site_meth))
```

Using Beta regression to test differences in global methylation as a function of sex, age, and genotype. None of the factors and interactions is significant.
```{r, global_meth_percent_betareg}
meta.gm <-  left_join(meta.gm, global_meth_perc, by="sample_id")

br.df <- meta.gm %>%
  mutate(genotype = recode(genotype, "CC"=0, "CT"=1, "TT"=2),
         sex = recode(sex, "F"=0, "M"=1),
         age = recode(age, "6"=0, "18"=1))
   
model = betareg(global_meth_perc ~ genotype + age + sex, data = br.df)
emmeans::joint_tests(model)

ggplot(meta.gm, aes(x=genotype, y=global_meth_perc, color=age)) +
  geom_point() +
  facet_wrap(~ sex)

```

### Correlation with homocysteine level

```{r hmc, out.width=c('50%', '50%')}
ggplot(meta.gm, aes(x = hmc, y = global_meth_counts)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(3,1.5)) +
  ggtitle("Counts")

ggplot(meta.gm, aes(x = hmc, y = global_meth_perc)) + 
  geom_point(aes(shape=sex, color=genotype, size=age)) +
  scale_size_discrete(range=c(3,1.5)) +
  ggtitle("Percent")
```


## Output
Uploading sample metadata with their global methylation values to [Synapse](https://www.synapse.org/#!Synapse:syn23573590/wiki/607402)
```{r output, results='hide'}
write_csv(meta.gm, file="output/RRBS/2_Exploratory_Analysis/RRBS_meta_globalmeth.csv")

# upload meta_global_meth to synapse
synLogin() 

file <- File("output/RRBS/2_Exploratory_Analysis/RRBS_meta_globalmeth.csv",
             contentType = "text/plain",
             description = "RRBS sample metadata with their global methylation values",
             parent = "syn25174486")

provenance <- Activity(used = c("syn25174542", "syn25174517"),
                       executed = "https://github.com/TheJacksonLaboratory/MTHFR_C667T/blob/master/RRBS_2_Exploratory_Analysis.Rmd")

file <- synStore(file, activity=provenance)
```

